A non-additive generalization of probability and \\ state count for all physical theories

### 2025-12-17 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: reviewing draft

\title{A non-additive generalization of measure theory and \\ probability theory for quantum mechanics and beyond}
\author{Gabriele Carcassi}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\author{Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

@@

\begin{abstract}
	We present a physically motivated generalization of probability theory that is suitable for classical mechanics, quantum mechanics and any future physical theory that allows a statistical description. The goal is to put the use of classical and quantum probability in a broader context, and show how the current mathematical structures are likely not suitable to solve the open problems in the foundations of physics. For the more mathematically inclined, we will point to areas where new math or generalization of established math are needed. For the more philosophically incline, we will point to areas where further conceptual work is needed.

@@

A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics\cite{chsh1969,fine1982}. One possible generalization is through quasiprobability,\cite{zachos2005,mandel1995,monchietti2025} which signed probability measures. However, given that one can non-negative quasiprobability representations of quantum spaces, the line between standard probability and quasiprobability has not been adequately established. Moreover, the measure that quantities the number of states in a region is also fundamental, as it is used both to define probability density and to define the notion of the entropy. Currently, a quantum analog of the count of states has not been properly identified.

\textbf{Article aim.} In this work we aim to fully explore the connection between physical theories and measure theoretic tools. Arguing from necessity, we want to find a minimal set of physical requirements for a state space of a physical system that allows us to directly connect to all the measure theoretic structures currently used in physics (i.e.~probability measures, Liouville measure to count states, quasiprobability, ...) and point to a straightforward generalization of such structures.\footnote{Note that we believe that the requirements we set, though necessary, are not sufficient to generalize the Lie algebra given by the classical Poisson brackets and the quantum commutators, meaning they cannot fully recover mechanics. Nonetheless, they are enough to recover measure theoretic structures, the topic of this paper.}

The starting point is that any physical law deals with experimentally reproducible relationships. This implies a physical theory must give us the space of statistical ensembles (i.e.~the collection of all possible preparation procedures) that the theory allows. As a basic requirement, ensembles must be experimentally well-defined (leading to a topological space), allow statistical mixing (leading to a convex structure) and quantify the variability of the elements of the ensembles (leading to an entropy function).\footnote{The main point of departure from Generalized Probabilistic Theories (GPTs) is the entropy requirement. We found that a convex set is not enough to fully characterize an ensemble space, especially in the infinite dimensional case.} Additionally, though not necessary, we require having enough experimentally measurable quantities to identify each ensemble (i.e.~there is a family of affine functions that identify each ensemble through their values). Mathematically, these premises lead us to assume that an ensemble space is a convex subset of a metrizable second countable locally convex topological vector space. The question is, then, what generalized measure theoretic tool can quantify the number of states (reducing to the Liouville measure in the classical case) or the number of possible outcomes of a measurement (reducing to the dimension of a subspace in the quantum case)? What measure theoretic tool can represent each ensemble? If we are to define a physically meaningful notion of probability density, both of these questions must be answered.

\textbf{Main finding.}  Choquet theory gives us a representation of ensembles in terms of convex combination of extreme points (i.e.~pure states). Classical spaces are those for which this representation is unique. Psuedo-probability measures are affine representations of the ensemble space, and different representation chosen based on the problem at hand. The probability measure finds a natural non-additive generalization in the fraction capacity $\frcap_{\ens}(A)$, which returns the highest possible fraction that an ensemble $\ens$ can be recovered from a convex combination of elements of $A$. The state counting measures finds a natural non-additive generalization in the state capacity $\capacity(A)$ which returns the highest exponential of the entropy (i.e.~entropic volume) achievable with a convex combination of elements of $A$.

@@

A final remark before we start. Too often works in the foundations of physics focus on a particular aspect without concern to a broader connection to all relevant mathematical, physical and philosophical aspects, such as information geometry, functional analysis, quantum logic and so on. However, such a complete work would be effectively overwhelming both to produce and to read. The main part of the paper, then, highlights the main conceptual results regarding measure theoretic representations, the appendix goes through the relevant mathematical detail while all the rest is left to our evolving collaborative open source work.\cite{aop-book}

@@

\textbf{Basic requirements for an ensemble space}. In our larger work, we strive to give all axioms and mathematical definitions solid grounding in physical requirements.\footnote{We call this approach Physical Mathematics, as starting from physical ideas recovers the mathematical structures, and only the mathematical structures, that are physically significant.} This way we can develop mathematical frameworks that are physically well founded. Specifically, we are developing a general theory of ensemble spaces, the full treatment of which is beyond the scope of this work. Here we limit ourselves to giving only the initial premises and the final mathematical characterization, with more details in the appendix and in our ongoing work.\cite{aop-book}

We start by assuming that physical laws deal with experimentally reproducible relationships, which implicitly requires the notion of statistical ensembles. Conceptually, a statistical ensemble is the collection of outputs, taken at once, of a reproducible preparation procedure for a physical system. In the classical case, this corresponds to a probability distribution over a discrete set of outcomes (e.g. a fair coin) or over phase space (e.g. the symplectic manifold representing position and momentum for a classical particle). In the quantum case, this corresponds to a mixed state (e.g. a positive semi-definite self-adjoint operator of trace one). A physical law, then, ultimately expresses relationships among ensembles. A minimal requirement for a physical theory, then, is to provide an ensemble space: the collection of all possible ensembles within that theory, together with a basic mathematical structure that all physical theories must possess.

An ensemble space $\Ens$ will need to satisfy three basic requirements. First, it needs to guarantee experimental verifiability. Mathematically, the ensemble space must be a $\mathsf{T}_0$ second countable topological space, where each open set represents an experimentally verifiable statement\cite{aop-book,kelly1996}.\footnote{A verifiable statement is an assertion for which there exists an experimental test that finishes successfully in finite time if and only if the statement is true.} Second, it needs to allow statistical mixtures: if $\ens[a],\ens[b] \in \Ens$ are ensembles, then $p \ens[a] + \bar{p} \ens[b]$ represents the statistical mixture where $\ens[a]$ is taken with probability $p$ and $\ens[b]$ with probability $\bar{p} = 1 - p$. Mathematically, this constrains the ensemble space to be a convex set.\footnote{Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i=1}^{\infty} p_i \ens_i$ converges in the space or not.} Third, an entropy function must be defined that characterizes the variability of the elements within an ensemble.

Additionally, we will require that ensembles can be identified through real valued statistical quantities.\footnote{It is an open question whether this is an additional assumption or it can be derived from the other requirements.} A statistical quantity represents the expectation of a real valued quantity, and is therefore an affine continuous function $F : \Ens \to \mathbb{R}$. That is, $F(p \ens[a] + \bar{p} \ens[b]) = p F(\ens[a]) + \bar{p} F(\ens[b])$. The last requirement is that there exists a family $F_i$ of statistical quantities that fully identify each ensemble. That is, $\ens[a] = \ens[b]$ if and only if $F_i(\ens[a]) = F_i(\ens[b])$ for all $i$.

>>>>

\title{A non-additive generalization of probability and \\ state count for all physical theories}
\author{Gabriele Carcassi}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\author{Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

@@

\begin{abstract}
	We present a physically motivated non-additive generalization of probability and state counting measures that is equally valid on all physical theories that allow a statistical description. In the classical case, these reduce to the standard probability and Liouville/counting measures, but retain useful properties in the general case. Both constructions are instances of a more general construction that may may be of more general interest. Moreover, the expectation for statistical variables is not recovered by standard non-additive integrals, and therefore new techniques will need to be developed.

@@

A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics\cite{chsh1969,fine1982}. The default approach is simply to view quantum states as collections of probability measures over different quantities. Another approach is to extend probability measures to signed measures\cite{zachos2005,mandel1995,monchietti2025}, though it is not clear what negative probability represent, particularly since non-negative quasiprobability representations of quantum spaces exist. Yet another approach is to replace the $\sigma$-algebra with a different structure (CITE some quantum logic, quantum measure theory ). The situation, then, is that we multiple approaches, each with its own strength and weakness, with no clear connection or straight forward physical justification. Moreover, none of them generalize the connection between entropy and the classical Liouville measure for counting states.

The aim of this paper is to present a simple, physically motivated structure in which all these approaches can be connected. Additionally, we show that a natural extension of probability and state counting measures exist under very general conditions.

The main findings are the following:
\begin{itemize}
	\item any physical theory needs to provide a space of statistical ensemble, which need to be a bounded convex subsets of a locally convex second countable topological vector space
	\item Choquet theory represents each ensemble $\ens$ with one or more probability measures $p_{\ens}$ on the set of extreme points $X$ (i.e. pure states)
	\item if this representation is unique (i.e. the ensemble space is a simplex) classical probability theory can be used
	\item the classical case corresponds to pure states being mutually exclusive
	\item quasiprobability approaches provide affine representions of ensembles in terms of a function space, whose points are unrelated to the extreme points
	\item the non-positivity of the measures is not key to quasiprobability approaches, and classical spaces admit quasiprobability as well
	\item the algebras used in quantum logic and quantum measure theory can always be understood as sub-algebras of the Borel algebra of ensembles
	\item a generalization of probability measures $p_{\ens}(A)$ emerges by asking the question ``what is the biggest fraction of a particular ensemble $\ens$ that can be expressed as a mixture of ensembles of $A$?''
	\item on the set of extreme points $X$, the generalized probability measure is the supremum of all Choquet representations, and becomes additive in the classical case
	\item a generalization of the state counting measure $\mu(A)$ emerges by asking ``what is the biggest spread that can be achieved with a mixture of ensembles of $A$?''
	\item  the generalized state counting measure reduces to the Liouville measure in the classical case, and to the dimensionality of subspaces in the quantum case
	\item to properly generalize probability theory, a non-additive notion of integral and derivative is needed which does not seem to match standard approaches.
\end{itemize}

A final remark before we start. Too often works in the foundations of physics focus on a particular aspect without concern to a broader connection to all relevant mathematical, physical and philosophical aspects, such as information geometry, functional analysis, quantum logic and so on. At this point in time, what is needed is a synthesis of many different idea. However, such work carried to completion would be effectively overwhelming both to produce and to read. Therefore, the main part of the paper can only highlight the main conceptual results regarding measure theoretic representations. In the appendix wit go through some of the relevant mathematical detail while all the rest is left to our evolving collaborative open source work.\cite{aop-book}

@@

@@

\textbf{Basic requirements for an ensemble space}. In our larger work, we strive to give all axioms and mathematical definitions solid grounding in physical requirements.\footnote{We call this approach Physical Mathematics, which stems from the observation that when physical objects are mapped to the right mathematical objects, the physical specification maps to the mathematical definitions.} Here we will only present a summary for the present investigation, leaving details to the appendix and our ongoing work.\cite{aop-book}

Since physical laws deal with experimentally reproducible relationships, they are inherently relationships between statistical ensembles. Therefore, any physical theory must define the space of all ensembles allowed within the theory. In the classical case, this corresponds to a probability distribution over a discrete set of outcomes (e.g. a fair coin) or over phase space (e.g. the symplectic manifold representing position and momentum for a classical particle). In the quantum case, this corresponds to a mixed state (e.g. a positive semi-definite self-adjoint operator of trace one).

In this work, we will take an ensemble space $\Ens$ to be a bounded convex subset of a locally convex metrizable second countable topological vector space endowed with an entropy function. The topology captures experimental verifiability as open sets represent experimentally verifiable statements\cite{aop-book,kelly1996}.\footnote{A verifiable statement is an assertion for which there exists an experimental test that finishes successfully in finite time if and only if the statement is true.} Convexity, and linearity more in general, comes from allowing statistical mixtures: if $\ens[a],\ens[b] \in \Ens$ are ensembles, then $p \ens[a] + \bar{p} \ens[b]$ represents the statistical mixture where $\ens[a]$ is taken with probability $p$ and $\ens[b]$ with probability $\bar{p} = 1 - p$.\footnote{Only finite mixtures are guaranteed, while the topology will decide whether an infinite mixture $\sum_{i=1}^{\infty} p_i \ens_i$ converges in the space or not.} Lastly, the entropy characterizes the variability of the elements within an ensemble. In this context, a statistical quantity, representing the expectation of a real valued quantity, is an affine continuous function $F : \Ens \to \mathbb{R}$. That is, $F(p \ens[a] + \bar{p} \ens[b]) = p F(\ens[a]) + \bar{p} F(\ens[b])$.
